### 2022-08-03 paper/localpsf.tex
minor edit to paper: appendix listed rational function order as l=1, when it is now l=2. Fixed.
fix missing references

We solve the inverse problem with an inexact Newton preconditioned conjugate gradient (PCG) scheme \cite{NOCEDAWRIGHT}, using $\basalfriction_0$ as the initial guess. The discretized parameter, $\bm{\basalfriction}$ is updated iteratively according to the formula
$
	\bm{\basalfriction}_{k+1} = \bm{\basalfriction}_k + \alpha_k \bm{\searchdir}_k,
$
where $\bm{\searchdir}_k$ approximately solves one of the following Newton linear systems,
\begin{equation}
	\label{eq:newton_system}
	\bm{H} \bm{\searchdir}_k = - \bm{g}_k \qquad \text{or} \qquad \bm{H}_\text{gn} \bm{\searchdir}_k = - \bm{g}_k,
\end{equation}
and the step length $\alpha_k \in \mathbb{R}$ is chosen by line search to satisfy the Armijo conditions \cite{ARMIJO}. Here $\bm{g}_k$ is the discretized gradient of $J$, $\bm{H}$ is the discretized Hessian, and $\bm{H}_\text{gn}$ is the discretized Gauss-Newton Hessian, and these derivatives are evaluated at $\bm{\basalfriction}_k$. 
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
We use $\bm{H}_\text{gn}$ for the first five iterations, and $\bm{H}$ for all subsequent iterations.  The Newton iteration terminates when $||\bm{g}_k||/||\bm{g}_0|| < 10^{-8}$. Systems \eqref{eq:newton_system} are solved inexactly using an inner PCG iteration, which terminates when the norm of the residual for the linear system is less than $\min \left(0.5, \sqrt{||\bm{g}_k||}\right)||\bm{g}_k||$. PCG only requires applying $\bm{H}$ (or $\bm{H}_\text{gn}$) to vectors. Each application of $\bm{H}$ (or $\bm{H}_\text{gn}$) to a vector requires solving one incremental forward and one incremental adjoint Stokes PDE of the form \eqref{Stokeseqn}, but with different right hand sides.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 impulse responses batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term to form a H-matrix approximation of $\bm{H}_\text{gn}$. Straightforward calculation shows that the discretized Hessian of the regularization term is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques. We factor the overall Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$, using fast H-matrix methods \cite{RONKRIEMANNEMAIL}, then and use the factorization as a preconditioner \cite{HLIBPRO}.

@@

Computing $\mathbf{\Pi}=\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by a high order rational function $\Pi_\ratord \approx \Pi$ which is defined as follows:
\begin{equation*}
	 \Pi_\ratord(\lambda) := \frac{1}{1 + T(\lambda)^{2^\ratord} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a}.
\end{equation*}
Here $\ratord$ is a positive integer; the larger $\ratord$ is, the better the approximation $\Pi \approx \Pi_\ratord$. 
%This is illustrated in Figure \ref{fig:spd_rat}. 
Replacing scalar operations with matrix operations, we have an approximation $\mathbf{\Pi}_\ratord \approx \mathbf{\Pi}$ given by
\begin{equation}
\label{eq:rational343}
\mathbf{\Pi}_\ratord := \Pi_\ratord\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^\ratord}\right)^{-1},
\end{equation}
where
$
\mathbf{T} := T(\mathbf{A}_H^{\text{sym}}) = \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
$
Using $\mathbf{\Pi}_\ratord$ instead of $\mathbf{\Pi}$ in \eqref{eq:I_Pi_abs} allows us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
This approximation procedure is as follows. First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. 
The method is relatively insensitive to $\gamma$ and numerically we find that the method works well for any $\gamma \in [1,2]$. We use $\gamma=1.5$ in our numerical results. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^\ratord}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using H-matrix multiplication, which is fast. We recommend using $\ratord=1$ or $2$ (we use $\ratord=1$ in our numerical results). Larger $\ratord$ is unnecessary and may make the method unstable. The approximate spectral projector $\mathbf{\Pi}_\ratord$ is formed from $\mathbf{T}^{2^\ratord}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\mathbf{\Pi}_\ratord\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. Note that for any $\ratord \ge 1$, $\Pi_\ratord(\lambda)> 1/2$ for $\lambda \in (a,b)$, so $\mathbf{A}_H^{\text{sym}+}$ is positive semi-definite\footnote{Some indefiniteness may be introduced by truncation error in the H-matrix operations, but the negative eigenvalues caused by this truncation error are typically extremely small.}. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced by truncation error in the H-matrix operations. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.

>>>>

We solve the inverse problem with an inexact Newton preconditioned conjugate gradient (PCG) scheme \cite[Section 7.1]{NocedalWright99}, using $\basalfriction_0$ as the initial guess. The discretized parameter, $\bm{\basalfriction}$ is updated iteratively according to the formula
$
	\bm{\basalfriction}_{k+1} = \bm{\basalfriction}_k + \alpha_k \bm{\searchdir}_k,
$
where $\bm{\searchdir}_k$ approximately solves one of the following Newton linear systems,
\begin{equation}
	\label{eq:newton_system}
	\bm{H} \bm{\searchdir}_k = - \bm{g}_k \qquad \text{or} \qquad \bm{H}_\text{gn} \bm{\searchdir}_k = - \bm{g}_k,
\end{equation}
and the step length $\alpha_k \in \mathbb{R}$ is chosen by line search to satisfy the Armijo conditions \cite[Section 3.1]{NocedalWright99}. Here $\bm{g}_k$ is the discretized gradient of $J$, $\bm{H}$ is the discretized Hessian, and $\bm{H}_\text{gn}$ is the discretized Gauss-Newton Hessian, and these derivatives are evaluated at $\bm{\basalfriction}_k$. 
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
We use $\bm{H}_\text{gn}$ for the first five iterations, and $\bm{H}$ for all subsequent iterations.  The Newton iteration terminates when $||\bm{g}_k||/||\bm{g}_0|| < 10^{-8}$. Systems \eqref{eq:newton_system} are solved inexactly using an inner PCG iteration, which terminates when the norm of the residual for the linear system is less than $\min \left(0.5, \sqrt{||\bm{g}_k||}\right)||\bm{g}_k||$. PCG only requires applying $\bm{H}$ (or $\bm{H}_\text{gn}$) to vectors. Each application of $\bm{H}$ (or $\bm{H}_\text{gn}$) to a vector requires solving one incremental forward and one incremental adjoint Stokes PDE of the form \eqref{Stokeseqn}, but with different right hand sides.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 impulse responses batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term to form a H-matrix approximation of $\bm{H}_\text{gn}$. Straightforward calculation shows that the discretized Hessian of the regularization term is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques. We factor the overall Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$, using fast H-matrix methods, then and use the factorization as a preconditioner.

@@

Computing $\mathbf{\Pi}=\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by a high order rational function $\Pi_\ratord \approx \Pi$ which is defined as follows:
\begin{equation*}
	 \Pi_\ratord(\lambda) := \frac{1}{1 + T(\lambda)^{2^\ratord} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a}.
\end{equation*}
Here $\ratord$ is a positive integer; the larger $\ratord$ is, the better the approximation $\Pi \approx \Pi_\ratord$. 
%This is illustrated in Figure \ref{fig:spd_rat}. 
Replacing scalar operations with matrix operations, we have an approximation $\mathbf{\Pi}_\ratord \approx \mathbf{\Pi}$ given by
\begin{equation}
\label{eq:rational343}
\mathbf{\Pi}_\ratord := \Pi_\ratord\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^\ratord}\right)^{-1},
\end{equation}
where
$
\mathbf{T} := T(\mathbf{A}_H^{\text{sym}}) = \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
$
Using $\mathbf{\Pi}_\ratord$ instead of $\mathbf{\Pi}$ in \eqref{eq:I_Pi_abs} allows us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
This approximation procedure is as follows. First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. 
The method is relatively insensitive to $\gamma$ and numerically we find that the method works well for any $\gamma \in [1,2]$. We use $\gamma=1.5$ in our numerical results. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^\ratord}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using H-matrix multiplication, which is fast. We recommend using $\ratord=1$ or $2$. Larger $\ratord$ is unnecessary and may make the method unstable. The approximate spectral projector $\mathbf{\Pi}_\ratord$ is formed from $\mathbf{T}^{2^\ratord}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\mathbf{\Pi}_\ratord\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. Note that for any $\ratord \ge 1$, $\Pi_\ratord(\lambda)> 1/2$ for $\lambda \in (a,b)$, so $\mathbf{A}_H^{\text{sym}+}$ is positive semi-definite\footnote{Some indefiniteness may be introduced by truncation error in the H-matrix operations, but the negative eigenvalues caused by this truncation error are typically extremely small.}. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced by truncation error in the H-matrix operations. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.
